Real-time modelling of a public transport network to improve arrival-time predictions

### 2018-02-07 realtime-filter/realtime-filter.Rnw
Some stuff about EKF update

That ensure that as $n$ goes to infinity, $\F_n(\bbeta, \bnu, \bpsi) \to \bnu$.
To do the same with the variance, we need to find a value for $\bQ$,
the variance of system noise, which is involved in the prediction of the state variance, $\bP$:
\begin{align}
  \label{eq:kf_varpred}
  \bP_{c|c-1} &= \F_n \bP_{c-1|c-1}\F_n^T + \bQ \nonumber \\
  \intertext{At equilibrium, we want $\bP_{c|c-1} = \bP_{c-1|c-1} = \bxi$,
  which we can also compute independently for each segment (since the off-diagonals will be 0):}
  \xi^m &= F'_n(\beta^m) \xi^m F'_n(\beta^m)^T + Q_{m,m} = (1-\psi^m)^{2n} \xi^m + Q_{m,m} \nonumber \\
  \intertext{which, solving for $Q$, gives us}
  Q_{m,m} &= \xi^m(1 - (1 - \psi^m)^{2n})
\end{align}

The last step is to choose a value of $\bpsi$.
However, any fixed value will cause the system to converge quickly when the difference is large;
instead, we want the convergence to be a function of the ratio of information held
in the prior and posterior states. 
For example, if we have a lot of buses traveling at the same speed along a segment, 
we will have a lot of information and want the state estimate to remain there
longer that if only one bus went through. 
So we will use $\psi^m = \phi \frac{P^m_{c-1|c-1}}{P^m_{c-1|c-1} + \xi^m}$.

\vspace{2em}
\hrule
Now that each vehicle is modelled by a sample of particles,
we can use the speed estimates of buses along segments to estimate
the speed along them, regardless of the route.
In future we can add correlation structure between adjacent segments.

Use Kalman filter because we have one parameter, speed, 
which we expect to be unimodal and approximately Gaussian
(some buses travel faster or slower).

Use weighted speed estimates as data,
and weighted variance as measurement error.

Want to find the state of all $M$ segments in the network,
% \begin{equation}
%   \label{eq:kf_state}
%   \bbeta = [\beta_1, \cdots, \beta_M]^T
% \end{equation}
which is the travel time along each segment,
along with the variance-covariance matrix $\Sigma$.
For now, we are assuming segments are independent,
so $\Sigma$ is a diagonal matrix.

We run the Kalman filter every time we run the particle filter,
since it has a low cost, allowing the most recent data to be incorporated.
For each vehicle $v$, we have $I_v$ particles which we use to
compute the mean distance into trip,
\begin{equation}
  \label{eq:weighted_mean}
  \bar x_v = \sum_{i=1}^{I_v} w_v^{(i)} x_v^{(i)}
\end{equation}
which we can use to estimate which segment the vehicle is in
(using the same method as in the initialisation step).

When the vehicle passes through a segment, $r_\ell$,
we set the travel time of the vehicle using the 
weighted mean and sample variance of the particles,
\begin{align}
  \label{eq:weighted_travel}
  \bar b_{v,\ell} &= \sum_{i=1}^{I_v} w_v^{(i)} b_{v\ell}^{(i)} \\
  \bar S_{v,\ell}^2 &= \sum_{i=1}^{I_v} w_v^{(i)}(b_{v\ell}^{(i)} - \bar b_{v\ell})
\end{align}

Having processed all vehicles, we move on to updating each \emph{segment},
$R_m$, $m = 1, \ldots, M$.
Define
\begin{equation*}
  V_m = \{v:\text{ vehicle $v$ passed through segment } R_m\}
\end{equation*}
and let $K_m$ be the length of $V_m$,
and $\ell(R_m)$ be the segment index of the given stop
(i.e., $\ell(r_\ell) = \ell$).
\begin{align}
  \label{eq:kf_data_estimates}
  \hat B_m &= \frac{1}{K_m} \sum_{v \in V_m} \bar b_{v,\ell(R_m)}  \\
  \hat Q_m &= \frac{1}{K_m} \sum_{v \in V_m} 
  \left(\bar b_{v\ell(R_m)}^2 + \bar S_{v\ell(R_m)}^2\right) - \hat B_m^2
\end{align}

Next is just the simple matter of applying the Kalman filter algorithm 
\cite{Anderson_2012} 
to the transport road network state,
using the data \mbox{$\hat {\mat B}_m = [\hat B_1, \ldots, \hat B_M]$} 
as the observation vector
and \mbox{$\hat {\mat Q}_k = \diag(\hat Q_1, \ldots, \hat Q_M)$},
as the measurement error matrix.
The transition matrix for the network state, $\mat F_c$,
is currently an identity matrix,
but in future we'll allow for the update to account for
changes in adjacent segments.
The Kalman filter prediction equations are
\begin{align}
  \bbeta_{c|c-1} &= \mat F_c\bbeta_{c-1|c-1} \\
  \Sigma_{c|c-1} &= \Sigma_{c-1|c-1} + \bW_c
\end{align}
where $\bW_c \sim \mathcal{N}(0, \delta_c\xi^2\eye_M)$, 
$\xi^2$ is the system noise parameter,
and $\delta_c = t_c - t_{c-1}$ is the time since the segment
was last updated.

The update equations simplify to
\begin{align}
  \bK_c &= \frac{\Sigma_{c|c-1}}{\Sigma_{c|c-1} + \hat {\mat Q}_c} \\
  \bbeta_{c|c} &= \bbeta_{c|c-1} + \bK_c(\hat {\mat B}_c - \bbeta_{c|c-1}) \\
  \Sigma_{c|c} &= (\eye_M - \bK_c) \Sigma_{c|c-1}
\end{align}
which make use of the segment means and variances computed in~(\ref{eq:kf_data_estimates}).

>>>>

That ensure that as $n$ goes to infinity, $\F_n(\bbeta, \bnu, \bpsi) \to \bnu$.
To do the same with the variance, we need to find a value for $\bQ$,
the variance of system noise, which is involved in the prediction of the state variance, $\bP$:
\begin{equation}
  \label{eq:kf_varpred}
  \bP_{c|c-1} = \F_n \bP_{c-1|c-1}\F_n^T + \bQ 
\end{equation}
At equilibrium, we want $\bP_{c|c-1} = \bP_{c-1|c-1} = \bxi$,
which we can also compute independently for each segment (since the off-diagonals will be 0):
\begin{align}
  \xi^m &= F'_n(\beta^m) \xi^m F'_n(\beta^m)^T + Q_{m,m}\\
    &= (1-\psi^m)^{2n} \xi^m + Q_{m,m} \nonumber \\
  \intertext{which, solving for $Q$, gives us}
  Q_{m,m} &= \xi^m(1 - (1 - \psi^m)^{2n})
\end{align}

The last step is to choose a value of $\bpsi$.
However, any fixed value will cause the system to converge quickly when the difference is large;
instead, we want the convergence to be a function of the ratio of information held
in the prior and posterior states. 
For example, if we have a lot of buses traveling at the same speed along a segment, 
we will have a lot of information and want the state estimate to remain there
longer that if only one bus went through. 
So we will use $\psi^m = \phi \frac{P^m_{c-1|c-1}}{P^m_{c-1|c-1} + \xi^m}$,
where $\phi$ is a tuning parameter.

Now that the prediction step has been defined,
we can consider how to update the state given a new observation;
but first, we need to define what an "observation" is.

Due to the nature of the application,
we will not be updating the state \emph{immediately} after a vehicle
traverses a road segment;
instead, we will complete the particle filter algorithm for all buses,
and then deal with the data obtained.

For each vehicle $v$, we have $I_v$ particles,
with weights $w_v^{(i)}, i=1\ldots I_v$, which we use to
estimate the vehicle's travel time through segment $r_\ell$
by using the weighted mean and sample variance:
\begin{align}
  \label{eq:weighted_travel}
  \bar b_{v,\ell} &= \sum_{i=1}^{I_v} w_v^{(i)} b_{v\ell}^{(i)} \\
  \bar S_{v,\ell}^2 &= \sum_{i=1}^{I_v} w_v^{(i)}(b_{v\ell}^{(i)} - \bar b_{v\ell})^2
\end{align}

Having processed all vehicles, we move on to updating each \emph{segment},
$R_m$, $m = 1, \ldots, M$.
Define
\begin{equation*}
  V_m = \{v:\text{ vehicle $v$ passed through segment } R_m\}
\end{equation*}
and let $|V_m|$ be the number of elements in $V_m$,
and $\ell(R_m)$ be the segment index of the given stop
(i.e., $\ell(r_\ell) = \ell$).
\begin{align}
  \label{eq:kf_data_estimates}
  \hat B_m &= \frac{1}{|V_m|} \sum_{v \in V_m} \bar b_{v,\ell(R_m)}  \\
  \hat E_m &= \frac{1}{|V_m|} \sum_{v \in V_m} 
  \left(\bar b_{v\ell(R_m)}^2 + \bar S_{v\ell(R_m)}^2\right) - \hat B_m^2
\end{align}

Next is just the simple matter of applying the Kalman filter algorithm 
\cite{Anderson_2012} 
to the transport road network state,
using the data \mbox{$\hat {\mat B}_m = [\hat B_1, \ldots, \hat B_M]$} 
as the observation vector
and \mbox{$\hat {\mat E}_k = \diag(\hat E_1, \ldots, \hat E_M)$},
as the measurement error matrix.

The update equations for the \gls{ekf} state are now
\begin{align}
  \bK_c &= \bP_{c|c-1}(\bP_{c|c-1} + \hat {\mat E}_c)^{-1} \\
  \bbeta_{c|c} &= \bbeta_{c|c-1} + \bK_c(\hat {\mat B}_c - \bbeta_{c|c-1}) \\
  \bP_{c|c} &= (\eye_M - \bK_c) \bP_{c|c-1}.
\end{align}
